---
title: "Bonytail (GIEL) Size-Tied Maturation Hazard — Growth-Track Sensitivity"
subtitle: "Comparing Fabens-increment (F1, F2), age-based VBGF (A1), and the baseline scalar-lpsi joint robust-design fits"
author: "Marsh & Associates"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-width: 9
    fig-height: 5.5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(coda)
library(knitr)
library(kableExtra)
library(patchwork)
theme_set(theme_bw(base_size = 12))

apply_table_style <- function(kbl_obj, header_above = NULL) {
  if (!knitr::is_html_output()) return(kbl_obj)
  out <- kableExtra::kable_styling(kbl_obj, bootstrap_options = c("striped", "hover", "condensed"),
                                     full_width = FALSE)
  if (!is.null(header_above)) out <- kableExtra::add_header_above(out, header_above)
  out
}

# Fit files (GIEL-only joint robust-design model, RUN_PONDS = c(3, 6, 7);
# pond_new 1 = IP2, 2 = IP5, 3 = IP6)
FIT_FILES <- c(F1 = "../data/BWRobustDesign_MCMC_giel_F1.RData",
               F2 = "../data/BWRobustDesign_MCMC_giel_F2.RData",
               A1 = "../data/BWRobustDesign_MCMC_giel_A1.RData")
BASELINE_FILE <- "../data/BWRobustDesign_MCMC_giel.RData"
HAZARD_FILES <- c(F1 = "../data/GIEL_MaturationHazard_F1.RData",
                   F2 = "../data/GIEL_MaturationHazard_F2.RData",
                   A1 = "../data/GIEL_MaturationHazard_A1.RData")

POND_LABELS <- c("IP2", "IP5", "IP6")   # pond_new 1:3
N_MOY <- 7L
FIRST_FY <- 2017L

TRACK_DESC <- c(
  F1 = "Fabens increment, IPCA-only",
  F2 = "Fabens increment, IPCA + Cibola",
  A1 = "Age-based VBGF (shared-K, release-anchored), IPCA-only"
)

load_env <- function(path) { e <- new.env(); load(path, envir = e); e }

envs <- lapply(FIT_FILES, load_env)
env_base <- load_env(BASELINE_FILE)
hazard_envs <- lapply(HAZARD_FILES, load_env)
```

## Background

This report compares three sensitivity fits of the joint two-stage robust-design
model for bonytail (GIEL) at the three IPCA ponds (IP2, IP5, IP6), each using a
different growth-model-derived **size-tied juvenile-to-adult maturation
hazard** in place of the model's original single scalar `lpsi[GIEL]` (a
species-level, size-blind, time-constant monthly logit-hazard), against that
original **baseline** fit. The three tracks differ only in which growth model
supplied the hazard lookup table (`psi_lookup`, 8 TL bins x 24 months);
everything else about the joint model and data build is identical across the
four fits.

| Track | Growth-model source | Fit file |
|---|---|---|
| F1 | `r TRACK_DESC["F1"]` | `data/BWRobustDesign_MCMC_giel_F1.RData` |
| F2 | `r TRACK_DESC["F2"]` | `data/BWRobustDesign_MCMC_giel_F2.RData` |
| A1 | `r TRACK_DESC["A1"]` | `data/BWRobustDesign_MCMC_giel_A1.RData` |
| baseline | scalar `lpsi[GIEL]`, no hazard table | `data/BWRobustDesign_MCMC_giel.RData` |

All four fits use the same joint model structure (3 chains, 30k iterations /
10k burn-in / thin 10) and the same underlying `Fish`/contact data; see
`AGENTS.md`'s "GIEL Size-Tied Maturation Hazard" section for the full design
rationale and implementation notes, and `model/GIEL_MaturationHazard_Build.R`
for how each track's `psi_lookup` table was derived from its source growth
model's posterior.

## 1. The three hazard schedules

Each track converts posterior growth-model draws into a monthly maturation
hazard for a juvenile bonytail as a function of its TL bin at tagging/capture
and months elapsed since then. The panels below show, for each track, (a) the
monthly hazard curve for three representative TL bins, and (b) the cumulative
probability of having reached 250 mm TL after 6 months (one Apr–Sep growing
season) as a function of starting TL bin — the version of this curve that is
actually discretized into the joint model's `psi_lookup`, not the smoother
continuous-TL curves reported elsewhere for these growth models.

```{r hazard-curves, fig.height=8}
hazard_long <- imap_dfr(hazard_envs, function(e, nm) {
  psi <- e$psi_lookup
  as_tibble(psi, rownames = "TL_bin") |>
    mutate(bin_mid = e$TL_bin_mid, track = nm) |>
    pivot_longer(starts_with("m"), names_to = "delta_t", values_to = "hazard") |>
    mutate(delta_t = as.integer(gsub("m", "", delta_t)))
})

# 3 representative bins: low, mid, high
rep_bins <- hazard_envs[[1]]$TL_bin_label[c(1, 4, 8)]

p_hazard <- hazard_long |>
  filter(TL_bin %in% rep_bins) |>
  ggplot(aes(x = delta_t, y = hazard, color = TL_bin)) +
  geom_line() +
  facet_wrap(~track, ncol = 3) +
  labs(x = "Months since tagging/capture", y = "Monthly maturation hazard",
       color = "TL bin (mm)",
       title = "Monthly hazard by track and starting size bin")

f_at_6mo <- imap_dfr(hazard_envs, function(e, nm) {
  tibble(track = nm, bin_mid = e$TL_bin_mid, F_6mo = e$F_mat[, 6])
})

p_curve <- f_at_6mo |>
  ggplot(aes(x = bin_mid, y = F_6mo, color = track)) +
  geom_line() + geom_point() +
  labs(x = "TL at tagging/capture (mm, bin midpoint)",
       y = "P(TL >= 250mm after 6 months)",
       title = "Cumulative probability of reaching 250mm after one growing season")

p_hazard / p_curve
```

The three tracks disagree most at small starting sizes and converge as TL
approaches the 250 mm threshold — the same divergence already documented for
each growth model's own smooth threshold curve (`AGENTS.md`, "GIEL Size-Tied
Maturation Hazard" section): F1 (Fabens, IPCA-only) predicts the slowest
maturation for small fish, F2 (Fabens, IPCA + Cibola) and A1 (age-based VBGF)
predict progressively faster maturation at low TL.

## 2. Known-alive constraint (`N_tag >= known alive`)

For every pond x month with at least one fish known alive (tagged, not yet
removed, and either scanned in that window or known alive from a later
contact), the posterior 2.5% quantile of tagged abundance
(`sum(N_tag[pond, t, 1:2])`, juvenile + adult) must be at or above the
directly-counted known-alive total.

```{r known-alive-check}
known_alive_check <- function(env) {
  Fish <- env$rd_Fish
  samp <- as.matrix(env$rd_samples)
  T_PRIM <- max(Fish$end_t, na.rm = TRUE)
  ntag_cols <- grep("^N_tag\\[", colnames(samp), value = TRUE)
  idx <- do.call(rbind, lapply(strsplit(gsub("N_tag\\[|\\]", "", ntag_cols), ", "),
                                function(x) as.integer(x)))
  colnames(idx) <- c("pond", "t", "stage")
  df_idx <- as_tibble(idx) |> mutate(col = ntag_cols)

  n_checked <- 0L
  violations <- tibble()
  for (j in 1:3) {
    for (t in 1:T_PRIM) {
      known_alive <- Fish |>
        filter(pond_new == j, entry_t <= t, end_t >= t,
               last_alive_t >= t, count_from <= t) |>
        nrow()
      if (known_alive == 0) next
      n_checked <- n_checked + 1L
      cols <- df_idx |> filter(pond == j, t == !!t) |> pull(col)
      if (length(cols) == 0) next
      post_sum <- rowSums(samp[, cols, drop = FALSE])
      q025 <- quantile(post_sum, 0.025)
      if (q025 < known_alive) {
        violations <- bind_rows(violations,
                                 tibble(pond = j, t = t, known_alive = known_alive,
                                        post_q025 = q025, post_mean = mean(post_sum)))
      }
    }
  }
  list(n_checked = n_checked, violations = violations)
}

ka_results <- lapply(envs, known_alive_check)

ka_table <- imap_dfr(ka_results, function(r, nm) {
  tibble(Track = nm, `Pond-months checked` = r$n_checked, Violations = nrow(r$violations))
})
kable(ka_table, caption = "Known-alive constraint: pond-months checked vs. violations") |>
  apply_table_style()
```

All three tracks pass with zero violations.

## 3. Convergence

Excluding the chronically-poor-mixing nuisance families (`N_tag`, `Tal`, `U`,
`U0`, `RecJ`, `r`, `Atot` — recruitment/pool quantities documented throughout
`AGENTS.md` as mixing poorly in essentially every GIEL/XYTE fit in this
project regardless of any other model change), the remaining core
hyperparameters (survival, detection, maturation, offsets, hyper-priors)
converge acceptably in all three tracks.

```{r convergence}
NUISANCE_FAMILIES <- c("N_tag", "Tal", "U", "U0", "RecJ", "r", "Atot")

convergence_summary <- function(env) {
  rhat <- env$rhat_all[, "Point est."]
  fam <- gsub("\\[.*", "", names(rhat))
  core <- rhat[!fam %in% NUISANCE_FAMILIES]
  nuis <- rhat[fam %in% NUISANCE_FAMILIES]
  tibble(n_core = length(core), max_rhat_core = max(core, na.rm = TRUE),
         n_core_over_1.1 = sum(core > 1.1, na.rm = TRUE),
         n_nuisance = length(nuis), max_rhat_nuisance = max(nuis, na.rm = TRUE),
         n_nuisance_over_1.1 = sum(nuis > 1.1, na.rm = TRUE))
}

conv_table <- imap_dfr(envs, ~ mutate(convergence_summary(.x), Track = .y, .before = 1))
kable(conv_table, digits = 3,
      col.names = c("Track", "n core", "Max Rhat (core)", "Core n > 1.1",
                    "n nuisance", "Max Rhat (nuisance)", "Nuisance n > 1.1"),
      caption = "Convergence by parameter family") |>
  apply_table_style()
```

Worst core-parameter Rhats are among `lphiA[·,·]` (season-adult-survival) and
`dE[·]` cells, all at or below ~1.05 — consistent with, and no worse than,
prior GIEL fits in this project. The elevated nuisance-family Rhats are the
same floating-point/mixing artifact documented throughout `AGENTS.md` for
`N_tag`/`Tal` (pinned by the known-alive constraint, so within-chain SD is
near zero and Rhat is not meaningful) and for recruitment nuisance
parameters, not problems introduced by the hazard mechanism.

## 4. `lpsi_adj[GIEL]` plausibility

`lpsi_adj[1]` is the logit-scale correction the model applies on top of each
track's raw growth-model-derived hazard lookup table. A value near 0 means
the robust-design contact/stage data agree with that track's schedule as-is;
a positive value means the data want **faster** maturation than the schedule
predicts, a negative value **slower**.

```{r lpsi-adj}
lpsi_adj_table <- imap_dfr(envs, function(e, nm) {
  s <- e$rd_summary
  if (!"lpsi_adj[1]" %in% rownames(s)) return(tibble())
  tibble(Track = nm, mean = s["lpsi_adj[1]", "mean"],
         lo = s["lpsi_adj[1]", "2.5%"], hi = s["lpsi_adj[1]", "97.5%"])
})
kable(lpsi_adj_table, digits = 3,
      col.names = c("Track", "Mean", "2.5%", "97.5%"),
      caption = "lpsi_adj[GIEL] posterior by track") |>
  apply_table_style()
```

All three are credibly nonzero (no CI includes 0), and the *direction*
differs by track: F1's positive offset means the contact data want
maturation faster than the Fabens-IPCA-only schedule predicts — consistent
with F1 giving the *slowest* raw schedule of the three (Section 1). F2 and
A1's negative offsets, of similar magnitude to each other, mean their faster
raw schedules need to be slowed down to match the data. This is a genuine
difference between tracks that mirrors the underlying divergence between the
Fabens-increment and age-based VBGF growth-model curves, not measurement
noise.

## 5. Core hyperparameter comparison

```{r core-hyperparameters}
CORE_PARS <- c("mu_phi[1]", "sigma_phi[1]", "dJ[1]",
               "dE[1]", "dE[2]", "dE[3]",
               "lpsi[1]", "lpsi_adj[1]",
               "lp[1, 1]", "lp[2, 1]", "lp[3, 1]",
               "lp[1, 2]", "lp[2, 2]", "lp[3, 2]",
               "g_eff[1]")

fmt_ci <- function(m, lo, hi) sprintf("%.3f (%.3f, %.3f)", m, lo, hi)

core_table <- imap_dfr(envs, function(e, nm) {
  s <- e$rd_summary
  pars <- intersect(CORE_PARS, rownames(s))
  tibble(param = pars, track = nm,
         val = fmt_ci(s[pars, "mean"], s[pars, "2.5%"], s[pars, "97.5%"]))
})
base_pars <- intersect(setdiff(CORE_PARS, "lpsi_adj[1]"), rownames(env_base$rd_summary))
core_table_base <- tibble(
  param = base_pars, track = "baseline",
  val = fmt_ci(env_base$rd_summary[base_pars, "mean"],
               env_base$rd_summary[base_pars, "2.5%"],
               env_base$rd_summary[base_pars, "97.5%"]))

core_wide <- bind_rows(core_table, core_table_base) |>
  pivot_wider(names_from = track, values_from = val) |>
  select(param, F1, F2, A1, baseline) |>
  arrange(match(param, CORE_PARS))

kable(core_wide, caption = "Core hyperparameters: mean (95% CI), F1 / F2 / A1 / baseline") |>
  apply_table_style()
```

Survival hyperparameters (`mu_phi`, `sigma_phi`, `dE[1:3]`) are essentially
identical across all three hazard tracks and the baseline (differences well
under 0.1 logit) — the size-tied maturation submodel is largely separable
from survival/die-off estimation, matching the pattern already documented
for the XYTE `a1` size-split submodel. The real movement is concentrated in
`lpsi` (the untagged-pool annual maturation rate, which loses information as
the tagged-fish pathway becomes size-tied) and juvenile detection
(`lp[·, 1]`), both of which shift modestly and roughly monotonically from F1
-> F2 -> A1 as the underlying schedule's implied maturation speed increases.
`dJ[1]` (juvenile-vs-adult survival offset) also drifts down from F1 to A1 as
more juvenile time gets reallocated by the faster schedules, but stays
positive in all three fits (juvenile survival modeled above adult on the
logit scale in every GIEL fit to date) — a separately-investigated,
IP2-detection-asymmetry-linked finding (see `AGENTS.md`), not an artifact of
this sensitivity sweep.

## 6. FY2022 (IP2) / FY2020 (IP5) event-adjusted realized annual survival

Both confirmed die-off events (IP2, Jan–Jul 2022, 7 of 12 FY2022 months
flagged; IP5, Jan–Sep 2020, 9 of 12 FY2020 months flagged; see
`data/BWDieOffEvents.csv` and `AGENTS.md`) are handled by a per-pond fixed
survival-reducing offset `dE[pond]` applied only in the flagged months, on
top of that pond-season's baseline `lphiA`. Realized annual survival over the
event season is therefore `plogis(lphiA)^(12 - nEvt) * plogis(lphiA + dE)^nEvt`.

```{r event-survival}
event_windows <- tibble(
  pond_label = c("IP2", "IP5"),
  pond = c(1L, 2L),        # pond_new index
  season = c(6L, 4L),      # FY2022, FY2020
  nEvt = c(7L, 9L)
)

realized_survival <- function(env, ew) {
  post <- as.matrix(env$rd_samples)
  pmap_dfr(ew, function(pond_label, pond, season, nEvt) {
    lphiA_col <- sprintf("lphiA[%d, %d]", pond, season)
    dE_col <- sprintf("dE[%d]", pond)
    if (!lphiA_col %in% colnames(post) || !dE_col %in% colnames(post)) return(tibble())
    lphiA <- post[, lphiA_col]; dE <- post[, dE_col]
    surv <- plogis(lphiA)^(12 - nEvt) * plogis(lphiA + dE)^nEvt
    no_adj <- plogis(lphiA)^12
    tibble(pond_label = pond_label, season = season, nEvt = nEvt,
           mean = mean(surv), lo = quantile(surv, 0.025), hi = quantile(surv, 0.975),
           no_adjustment_mean = mean(no_adj))
  })
}

event_table <- imap_dfr(envs, function(e, nm) {
  mutate(realized_survival(e, event_windows), track = nm, .before = 1)
})

event_display <- event_table |>
  mutate(FY = if_else(pond_label == "IP2", 2022L, 2020L),
         `Event-adjusted (mean, 95% CI)` = fmt_ci(mean, lo, hi),
         `No-event-adjustment comparison` = round(no_adjustment_mean, 3)) |>
  select(Track = track, Pond = pond_label, FY, `Event-adjusted (mean, 95% CI)`,
         `No-event-adjustment comparison`)

kable(event_display, caption = "Event-adjusted realized annual survival by track") |>
  apply_table_style()
```

```{r event-survival-plot, fig.height=3.5}
event_table |>
  mutate(FY = if_else(pond_label == "IP2", "IP2, FY2022", "IP5, FY2020")) |>
  ggplot(aes(x = track, y = mean, ymin = lo, ymax = hi)) +
  geom_pointrange() +
  facet_wrap(~FY, scales = "free_y") +
  labs(x = "Track", y = "Event-adjusted realized annual survival",
       title = "Event-adjusted survival is stable across maturation-hazard tracks")
```

All three tracks give essentially the same event-adjusted realized survival
for both die-off seasons — confirming that the maturation-hazard track
choice does not materially affect the die-off (`dE`) estimates or their
downstream realized-survival implications, consistent with Section 5's
finding that the tracks differ mainly in maturation-timing quantities, not
survival.

## 7. Conclusion

All three growth-track sensitivity fits (F1, F2, A1) pass the known-alive
constraint with zero violations and converge acceptably (core-parameter Rhat
at or below ~1.05 in every fit). Survival and die-off estimates are robust to
the choice of maturation-hazard track — the size-tied maturation submodel is
largely separable from the rest of the joint model, mirroring the pattern
already seen for XYTE's density-dependent size-split submodel.

The substantive difference between tracks is the sign and magnitude of
`lpsi_adj[GIEL]`: F1 (the slowest raw schedule) is corrected faster by the
data, while F2 and A1 (faster raw schedules) are corrected slower by a
similar amount. This reflects the same underlying divergence between the
Fabens-increment and age-based VBGF growth-model curves already flagged as an
open reconciliation item in `AGENTS.md` and the design plan file. No track is
disqualified by this comparison; choosing among them, reconciling the two
growth models directly, or reporting all three as a sensitivity range on
maturation timing remains a modeling decision for BK. Because survival
estimates are unaffected by this choice, that decision can be made
independently of any of the survival/die-off conclusions already reported
elsewhere for this model.
