---
title: "Backwater Stocking and Capture Event Summary"
subtitle: "Backwater Genetics Study — Yuma Cove and Imperial Ponds"
author: "B. Kesner"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
    code-fold: true
  docx:
    toc: true
    toc-depth: 2
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

source("LabFunctions.R")
packages(dplyr)
packages(lubridate)
packages(knitr)
packages(flextable)

load("data/ReportingData.RData")

month_levels <- c("Jan","Feb","Mar","Apr","May","Jun",
                  "Jul","Aug","Sep","Oct","Nov","Dec")

# ── Stocking summary ──────────────────────────────────────────────────────────
# One row per fish per stocking event; count distinct PITIndex by BW / year / month

StockingSummary <- StudyBWNFWG |>
  filter(event == "stocking") |>
  mutate(
    MonthName = factor(month.abb[CollectionMonth], levels = month_levels)
  ) |>
  group_by(Backwater, CollectionYear, CollectionMonth, MonthName) |>
  summarise(
    Stocked  = n_distinct(PITIndex),
    Males    = sum(sex == "M",                na.rm = TRUE),
    Females  = sum(sex == "F",                na.rm = TRUE),
    Unknown  = sum(!sex %in% c("M", "F"),     na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(Backwater, CollectionYear, CollectionMonth)

# ── Capture summary ───────────────────────────────────────────────────────────
# Tagging status : FirstRecord == 1 → Newly Tagged; 0 → Recaptured
# Disposition    : Transfer == 0 & Mortality == 0 → Released
#                  Transfer == 1                  → Transferred
#                  Mortality == 1                  → Mortality

CaptureSummary <- StudyBWCaptureUniques |>
  mutate(
    MonthName = factor(month.abb[CollectionMonth], levels = month_levels)
  ) |>
  group_by(Backwater, CollectionYear, CollectionMonth, MonthName) |>
  summarise(
    `Newly Tagged` = sum(FirstRecord == 1),
    Recaptured   = sum(FirstRecord == 0),
    Released     = sum(Transfer == 0 & Mortality == 0),
    ReleasedNew  = sum(Transfer == 0 & Mortality == 0 & FirstRecord == 1),
    Transferred  = sum(Transfer == 1),
    .groups = "drop"
  ) |>
  arrange(Backwater, CollectionYear, CollectionMonth)

backwaters <- sort(unique(c(StockingSummary$Backwater,
                             CaptureSummary$Backwater)))
```

## Overview

This document summarises all stocking and capture events recorded in each study
backwater by year and calendar month. Each section covers one location.

**Stocking tables** show the number of PIT-tagged fish stocked in each
year / month.

**Capture tables** break down each year / month by:

- *Tagging status* — fish captured for the first time in this backwater
  (Newly Tagged) versus fish captured on a previous occasion (Recaptured).
- *Disposition* — fish returned to the backwater (Released) or removed and
  stocked elsewhere (Transferred).

---

```{r}
#| label: location-tables
#| results: asis

for (bw in backwaters) {

  # Section header
  cat("\n\n## ", bw, "\n\n", sep = "")

  # ── Stockings ──────────────────────────────────────────────────────────────
  stk <- StockingSummary |>
    filter(Backwater == bw) |>
    select(Year = CollectionYear, Month = MonthName, Stocked, Males, Females, Unknown) |>
    mutate(Year = as.character(Year))

  cat("### Stockings\n\n")

  if (nrow(stk) == 0) {
    cat("*No stocking records for this location.*\n\n")
  } else {
    stk |>
      flextable() |>
      add_header_row(values = c("", "Sex"), colwidths = c(3, 3)) |>
      theme_booktabs() |>
      bold(part = "header") |>
      align(align = "center", part = "header") |>
      align(j = "Month", align = "left", part = "body") |>
      align(j = c("Year", "Stocked", "Males", "Females", "Unknown"),
            align = "right", part = "body") |>
      colformat_num(j = c("Stocked", "Males", "Females", "Unknown"), big.mark = ",") |>
      bg(i = which(seq_len(nrow(stk)) %% 2 == 0), bg = "#F2F2F2", part = "body") |>
      autofit() |>
      flextable_to_rmd()
    cat("\n\n")
  }

  # ── Captures ───────────────────────────────────────────────────────────────
  cap <- CaptureSummary |>
    filter(Backwater == bw) |>
    mutate(
      `Released (Newly Tagged)` = paste0(Released, " (", ReleasedNew, ")"),
      Year = as.character(CollectionYear)
    ) |>
    select(Year, Month = MonthName,
           `Newly Tagged`, Recaptured,
           `Released (Newly Tagged)`, Transferred)

  cat("### Captures\n\n")

  if (nrow(cap) == 0) {
    cat("*No capture records for this location.*\n\n")
  } else {
    cap |>
      flextable() |>
      add_header_row(
        values    = c("", "Tagging Status", "Disposition"),
        colwidths = c(2, 2, 2)
      ) |>
      theme_booktabs() |>
      bold(part = "header") |>
      align(align = "center", part = "header") |>
      align(j = "Month", align = "left", part = "body") |>
      align(j = c("Year", "Newly Tagged", "Recaptured",
                  "Released (Newly Tagged)", "Transferred"),
            align = "right", part = "body") |>
      colformat_num(j = c("Newly Tagged", "Recaptured", "Transferred"),
                    big.mark = ",") |>
      bg(i = which(seq_len(nrow(cap)) %% 2 == 0), bg = "#F2F2F2", part = "body") |>
      autofit() |>
      flextable_to_rmd()
    cat("\n\n")
  }
}
```
